---
title: "a3Read In Files"
author: "Henry Kanengiser"
date: "2022-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(tidycensus)
library(sf)



# DIRECTORIES 
# (necessary because there are many files in this folder and I needed to use subfolders)
wd <- substr(getwd(), 1, nchar(getwd())-3)
dir <- paste0(wd, "csv")
dir2 <- paste0(wd, "dat/censustract")


# FILE NAMES
censusfil <- "censusacstract.shp"
nychafil <- "nycha_bbls.csv"
rentstab <- "rentstabunits.csv"


```

The **purpose** of this .Rmd file is to read in and do initial checking on the main tabular data used in this project. This includes data from NYC DCP, the Census, and from [nycdb](https://github.com/nycdb/nycdb), a collection of ready-to-use NYC housing data. At the end of this program, permanent cleaned files are saved, which will be joined and restructured in a later program.

This is the first of several programs used for this project. The second will be the summary step program (a3ss), where I consolidate and restructure these dataset. The thirs program will be the analysis program (a3an), where I conduct analyses and produce figures based on those analyses.

# Census data

I use Census data to describe demographic characteristics of the census tracts and block groups in the flood zone

```{r}

# Pull the variables from the ACS and the decennial census to identify what will be helpful for this analysis

# # HK commented out because we don't need to print these for the code to run
# cen20vars <- load_variables(2020, "pl", cache = TRUE)
acs20vars <- load_variables(2020, "acs5", cache = TRUE)

```

The census values will provide race/ethnicity information. The variables we want include: P2_001N through P2_011N
P2_001N (total population)
P2_002N (Hispanic or Latino)
P2_003N (not Hispanic or Latino)
P2_004N (not Hispanic or Latino: 1 race)
P2_005N (...: white alone)
P2_006N (...: Black or African American alone)
P2_007N (...: American Indian or Alaska Native alone)
P2_008N (...: Asian alone)
P2_009N (...: Hawaiian or Pacific Islander alone)
P2_010N (...: other race alone)
P2_011N (...: 2+ races)

The ACS variables are more complicated to hunt down. Median hh income comes from the following variables:
**Block group**
B19013_001 (median hh income in past 12 months)
table B19001 breaks down hh income by $5,000 categories (block group ONLY)
**Census tract**
B25119_001 (Median household income in the past 12 months)
B25119_002 (Med HH inc - Owner occupied)
B25119_003 (Med HH inc - Renter occupied)
B09019_003 (Total households)
B07013_002 (householder living in owner-occupied unit - tract)
B07013_003 (householder living in renter-occupied unit - tract)
More info about this term [here](https://www.census.gov/quickfacts/fact/note/US/HSG445220)

Estimated population living in owner/renter-occupied units:
**Block group**
B25008_001 (total)
B25008_002 (owner occupied)
B25008_003 (renter occupied)
**Census tracts**
B25026_001 (total)
B25026_002 (owner occupied)
B25026_009 (renter occupied)

Now, read in these datasets from tidycensus

## Decennial census
```{r}

# Filter the file down to just NYC
#--census tract
census_pop_tract <- get_decennial(
  geography = "tract",
  variables = c(
    poptot    = "P2_001N",
    pophis    = "P2_002N",
    popnhis   = "P2_003N",
    popnh1r   = "P2_004N",
    popwhite  = "P2_005N",
    popblack  = "P2_006N",
    popamind  = "P2_007N",
    popasian  = "P2_008N",
    poppacil  = "P2_009N",
    popoth1r  = "P2_010N",
    popnhis2r = "P2_011N"
  ),
  state = "New York",
  geometry = TRUE,
  year = 2020
) %>%
  clean_names() %>%
  filter(grepl("Bronx|Kings|New York County|Queens|Richmond", name)) %>%
  pivot_wider(id_cols = c("geoid", "name", "geometry"),
              names_from = "variable",
              values_from = "value")

# Check variable creation against the data.census.gov website
# https://data.census.gov/table?q=Race+and+Ethnicity&g=0500000US36005,36047,36061,36081,36085

census_pop_tract %>%
  st_drop_geometry() %>%
  mutate(borough = case_when(
    grepl("New York County", name) ~ "New York",
    grepl("Bronx", name) ~ "Bronx",
    grepl("Kings", name) ~ "Kings",
    grepl("Queens", name) ~ "Queens",
    grepl("Richmond", name) ~ "Richmond",)) %>%
  group_by(borough) %>%
  summarise(
    poptot = sum(poptot, na.rm = TRUE),
    pophis = sum(pophis, na.rm = TRUE),
    popwhite = sum(popwhite, na.rm = TRUE), 
    popblack = sum(popblack, na.rm = TRUE),
    popasian = sum(popasian, na.rm = TRUE)
  )

```

## ACS data

### Block group
```{r}

acs_block <- get_acs(
  geography = "block group",
  variables = c(
    med_hh_inc_bg = "B19013_001",
    pop_total_bg = "B25008_001",
    pop_owner_bg = "B25008_002",
    pop_renter_bg = "B25008_003"
  ),
  year = 2020,
  state = "New York",
  geometry = TRUE,
  survey = "acs5"
) %>%
  clean_names() %>%
  filter(grepl("Bronx|Kings|New York County|Queens|Richmond", name)) %>%
  mutate(moe_rate = moe/estimate) 

#Look at moe_rates (note that there are many beyond the range of 0-1)
acs_block %>%
  ggplot() + 
  aes(x = moe_rate) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0,1)) + 
  facet_wrap(~variable)

```


### Census tract
```{r}

acs_tract <- get_acs(
  geography = "tract",
  variables = c(
    mhhinc    = "B25119_001",
    mhhinco = "B25119_002",
    mhhincr = "B25119_003",
    pophoh  = "B09019_003",
    poptotal  = "B25026_001",
    popowner  = "B25026_002",
    poprenter = "B25026_009"
  ),
  year = 2020,
  state = "New York",
  geometry = TRUE,
  survey = "acs5"
) %>%
  clean_names() %>%
  filter(grepl("Bronx|Kings|New York County|Queens|Richmond", name)) %>%
  mutate(moerate = moe/estimate) 


acs_tract %>%
  ggplot() + 
  aes(x = moerate) + 
  geom_histogram() + 
  scale_x_continuous(limits = c(0,1)) + 
  facet_wrap(~variable)

```

Because the MOE's are considerably lower for census tracts than for block groups, we will use tracts going forward

Restructure the ACS tract data to be ready to merge

```{r}

acs_tract2 <- acs_tract %>%
  pivot_wider(id_cols = c("geoid", "name", "geometry"),
              names_from = "variable",
              names_glue = "{variable}{.value}",
              values_from = c("estimate", "moe", "moerate")) %>%
  #shorten the estimate varaible names
  rename(mhhinc = mhhincestimate,
         mhhinco = mhhincoestimate,
         mhhincr = mhhincrestimate,
         pophoh = pophohestimate,
         poptotal = poptotalestimate,
         popowner = popownerestimate,
         poprenter = poprenterestimate)

glimpse(acs_tract2)

```

## Join census and ACS data (census tract)

Join the ACS and Census data into one wide dataset that can be added to the MapPLUTO data in the next program.
Note that we need to recreate the borough-census tract combo that is used in PLUTO
-   7 digits long (1 digit for boro code, followed by 6 digits for census tract)
-   last two digits are for after the decimal point. if there's no decimal then pad with two zeroes

```{r}

census_acs_tract <- acs_tract2 %>% 
  #remove duplicate geometry
  st_drop_geometry() %>%
  select(-geoid) %>%
  #join by census tract name
  left_join(census_pop_tract, by = "name") %>%
  mutate(borough = case_when(
    grepl("New York County", name) ~ "1",
    grepl("Bronx", name) ~ "2",
    grepl("Kings", name) ~ "3",
    grepl("Queens", name) ~ "4",
    grepl("Richmond", name) ~ "5",
    TRUE ~ NA_character_
  ),
  #extract the census tract number
  census_tract_raw = gsub(".*Census Tract (.*?), .*", "\\1", name),
  #pull pre-dot part
  census_tract_predot = sub("\\..*", "", census_tract_raw),
  census_tract_postdot = ifelse(
    grepl("\\.", census_tract_raw), sub(".*\\.", "", census_tract_raw), "00"
    ),
  bct2020 = paste0(borough, str_pad(census_tract_predot, width = 4, side = "left", pad = "0"), census_tract_postdot))
  
# Check creation of each of these variables to make sure it fits the MapPLUTO convention
census_acs_tract %>%
  select(name, borough, census_tract_raw, census_tract_predot, census_tract_postdot, bct2020) %>%
  slice_sample(n = 100)

```

Prep to save it as a shapefile
```{r}
as.data.frame(
names(census_acs_tract),
substr(names(census_acs_tract), 1, 10)) %>%
  mutate(nchar = nchar(`names(census_acs_tract)`)) %>%
  arrange(desc(nchar))

# drop some of the longer variables that we no longer need

census_acs_tract2 <- census_acs_tract %>%
  select(-census_tract_postdot, 
         -census_tract_predot, 
         -census_tract_raw, 
         -contains("moerate"))

glimpse(census_acs_tract2)

# check if any important vars are still being cut off (no)
as.data.frame(
names(census_acs_tract2),
substr(names(census_acs_tract2), 1, 10)) %>%
  mutate(nchar = nchar(`names(census_acs_tract2)`)) %>%
  arrange(desc(nchar))

```



# NYCDB datasets

## NYCHA bbls
```{r}
url <- "https://raw.githubusercontent.com/JustFixNYC/nycha-scraper/098bd8232bee2cd59266bf278e26e32bd0cd7df1/Block-and-Lot-Guide-08272018.csv"

nycha_bbls_raw <- read.csv(url) %>% clean_names()
glimpse(nycha_bbls_raw)

# missing bbl, need to create it from borough block and lot

# check that block and lot are never more than 5 and 4 numbers long
nycha_bbls_raw %>%
  summarise(l_block = max(block),
            l_lot = max(lot))


nycha_bbls2 <- nycha_bbls_raw %>%
  mutate(borocode = case_when(
    borough == "MANHATTAN" ~ "1",
    borough == "BRONX" ~ "2",
    borough == "BROOKLYN" ~ "3",
    borough == "QUEENS" ~ "4",
    borough == "STATEN ISLAND" ~ "5",
    TRUE ~ "error"
  )) %>%
  mutate(bbl = paste0(
    borocode,
    str_pad(block, width = 5, side = "lef", pad = "0"),
    str_pad(lot, width = 4, side = "left", pad = "0")
  )) 

# spot check of var creation
nycha_bbls2 %>%
  select(borough, borocode, block, lot, bbl) %>%
  slice_sample(n = 100)



```

The dataset seems too long for the number of NYCHA developments, investigate duplicates in BBL and decide what to do
```{r}

nycha_bbls2 %>%
  group_by(bbl) %>%
  summarise(nrows = n()) %>%
  count(nrows)

# make sure that only one development name is associated with each BBL, then deduplicate
nycha_bbls2 %>%
  group_by(bbl) %>%
  mutate(namesame = ifelse(first(development) == last(development), 1, 0)) %>%
  ungroup() %>%
  filter(namesame == 0)
  
# count(namesame)

# these differences are understandable (BETANCES V vs BETANCES VI are the same development)
```


Since we are only interested in identifying the NYCHA developments associated with each BBL, we can deduplicate this dataset to the BBL level.

```{r}
# deduplicate, keeping the row with the most common development name
nycha_bbls3 <- nycha_bbls2 %>%
  left_join(
    nycha_bbls2 %>% count(bbl, development, name = "ndevelopment"),
    by = c("bbl", "development")
  )


nycha_bbls4 <- nycha_bbls3 %>%
  arrange(desc(ndevelopment)) %>%
  group_by(bbl) %>%
  distinct(bbl, .keep_all = TRUE) %>%
  select(-ndevelopment)

nycha_bbls4 %>% head(n=10)

```



## rent stabilized units per bbl

Note that there are two different files that I have to stack to get comprehensive data on # rent stabilize units dating back to 2007
```{r}

#older data (2007-2017)
url <- "https://taxbillsnyc.s3.amazonaws.com/joined.csv"

rentstabold <- read.csv(url) %>% clean_names()

glimpse(rentstabold)
```

```{r}
#newer data (2018-2020)
url <- "https://s3.amazonaws.com/justfix-data/rentstab_counts_from_doffer_2020.csv"

rentstabnew <- read.csv(url) %>% clean_names()

glimpse(rentstabnew)


```

Now, join the two rentstab files together into one

```{r}
rentstaball <- full_join(
  mutate(rentstabold, in_old = 1),
  mutate(rentstabnew, in_new = 1),
  by = "ucbbl"
)

# check success of the merge (do most rows line up?)
rentstaball %>% count(in_old, in_new)

# yes, but there are a substantial number that don't, look into why

# did the buildings on the old file have 0 stabilized units in the last year?
rentstaball %>%
  filter(is.na(in_new)) %>%
  mutate(none2017 = is.na(x2017uc)) %>%
  count(none2017)
# mostly (10%)

# i think i'll likely end up using the most recent year of data anyway, so this is fine

```


# Save permanent files
```{r}

## HK Commented out to avoid re-saving these each time I run the program. 
## Uncomment to save

# # Census Data
st_write(census_acs_tract2, dsn = file.path(dir2, censusfil), delete_dsn = TRUE)

# # NYCHA BBLs
# write_csv(nycha_bbls4, file = file.path(dir,nychafil))
# 
# # Rent stabilized units
# write_csv(rentstaball, file = file.path(dir, rentstab))

```



